Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-08 Manuscript/manuscript.tex

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
\todo{JOE: following doesn't sound right to me. The reasoning is weird.}
Since direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple model systems can be then
used to understand the role of ions in complex biological systems.

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered a higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{chowdhary13} by implicitly including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
\todo{JOE: It would be a good idea to be consistent in using ECC resp. MDEC. I think that here shuld be MDEC as it cites Stuchebrukhov.}
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids - but there the factor is not 0.5.}
However, the present concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11}.
%but significantly more simple than the explicit inclusion of electronic
%polarizability \cite{chowdhary13} both for derivation and application.

@@

In this work, we develop a phospholipid model for classical MD simulations that
accurately describes the lipid head group response to varying concentrations
of monovalent and divalent cations. This is a biologically highly relevant membrane feature, which is
poorly reproduced by currently available models which can affect not only on membrane properties
in the presence of ions but also modulates the interaction with charged moieties in the surface.
Importantly, this response from simulations can be accurately
compared against experimental NMR data \cite{akutsu81,altenbach84,scherer89},
as discussed in Ref. \citenum{catte16} in section \ref{section:electrometer}.
To this end, we empirically explore the scaling factor parameter space, $f_q \in [0.75, 1.0]$ for the Lipid14~\cite{dickson14} force field.
We selected this force field as a starting point because its response to bound ions was apparently the most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding
\todo{Do we really need to scale charges in non charged methylens in the choline and glycerol backbones? why? Does it really make a difference?
  JOE: Cannot answer this without actually doing it. The selection was a matter of choice -- in the beginnig we argued with Samuli about this.
  I said that we shuld select the polar atoms in this selection and leave out the rest as ECC would mean little difference to them. Samuli
  then argued in a way that if it makes no difference, why not apply ECC onto such atoms too? Hence we applied ECC on all of the atoms.}.
However, we do not modify the hydrocarbon chain parameters, because they
do not come in contact with salt ions and are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in most lipids models, including Lipid14 \cite{ollila16}.
This is in contrasts with the behavior in glycerol backbone and head group order
parameters which call for improvements in all available lipid models~\cite{botan15}.

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
The simulated systems consisted of POPC bilayer in pure water
or in varying salt concentrations.
In particular, the periodic orthorhombic simulation box contained 
128 POPC molecules and approximately 50~water molecules per each lipid.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
presented in Supporting Information (SI). 
We used the ECC-ions model for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion models by Dang~\cite{smith94,chang1999,dang2006}, and by \AA{}qvist~\cite{aqvist90}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
  this is done in Amber by default. If this is correct, the information in the
  table do not match with this sentence. Based on semi-extensive testing I made few years
ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\begin{table}[btp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 100~ns  \\
    simulation time     & 200~ns  \\
    temperature         & 313~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
  \todo{This could be moved to SI. Only simulation lengths needs to be mentioned in the main paper.}
\end{table}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: Chemical structure of POPC with definitions of different order parameters calculated.
    Middle: Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.
    Bottom: X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    } 
  \todo{X-label misses glycerol region. JOE: glycerol belongs to the headgroup part, but, yes, it wouldn't hurt to write it explicitly. SAMULI: I would maybe remove the whole text, tell the information in caption and increase the order parameter figure size in z-direction.} \\
  \todo{Also acyl chain points should correspond the legends and maybe have the same size as HG and glycerol.} \\
  \todo{Legends would be more clear without multiple points.}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-} and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration of the ions in water is 350~mM in all systems
    presented here. 
    }
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{The current data for Dang simulation seems to contain more ions than others.}
\end{figure}

@@

>>>>

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple model systems can be then
used to understand the role of ions in complex biological systems.

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered a higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{chowdhary13} by implicitly including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids - but there the factor is not 0.5.}
However, the present concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11}.
%but significantly more simple than the explicit inclusion of electronic
%polarizability \cite{chowdhary13} both for derivation and application.

@@

In this work, we develop a phospholipid model for classical MD simulations that
accurately describes the lipid head group response to varying concentrations
of monovalent and divalent cations. This is a biologically highly relevant membrane feature, which is
poorly reproduced by currently available models which can affect not only on membrane properties
in the presence of ions but also modulates the interaction with charged moieties in the surface.
Importantly, this response from simulations can be accurately
compared against experimental NMR data \cite{akutsu81,altenbach84,scherer89},
as discussed in Ref. \citenum{catte16} in section \ref{section:electrometer}.
To this end, we empirically explore the scaling factor parameter space, $f_q \in [0.75, 1.0]$ for the Lipid14~\cite{dickson14} force field.
We selected this force field as a starting point because its response to bound ions was apparently the most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding
However, we do not modify the hydrocarbon chain parameters, because they
do not come in contact with salt ions and are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in most lipids models, including Lipid14 \cite{ollila16}.
This is in contrasts with the behavior in glycerol backbone and head group order
parameters which call for improvements in all available lipid models~\cite{botan15}.

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
The simulated systems consisted of POPC bilayer in pure water
or in varying salt concentrations.
In particular, the periodic orthorhombic simulation box contained 
128 POPC molecules and approximately 50~water molecules per each lipid.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
presented in Supporting Information (SI). 
We used the ECC-ions model for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion models by Dang~\cite{smith94,chang1999,dang2006}, and by \AA{}qvist~\cite{aqvist90}. 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: Chemical structure of POPC with definitions of different order parameters calculated.
    Middle: Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  
    from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.
    Error bars are approximately the size of the points for head group order parameters.
    Point size for the order parameters of tails are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    } 
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-} and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration of the ions in water is 350~mM in all systems
    presented here. 
    }
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
\end{figure}

@@

\begin{table}[btp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 100~ns  \\
    simulation time     & 200~ns  \\
    temperature         & 313~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
\end{table}
